---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%"
)
```

# vcglR


The Victorian gambling regulator publishes gaming machine data by venue and local government area [on its website.](https://www.vgccc.vic.gov.au/resources/information-and-data/expenditure-data)

Released twice a year for venue data and monthly for LGA data, these datasets include gaming expenditure at each gaming venue/local area, and the number of gaming machines it has.

The data is published as `.xlsx` data. Historical data is also available.

The goal of vcglR is to make Victorian gambling data available in tidy format. The name is taken from the abbreviation of the gambling regulators former name.

The `vcglR` package reads the `.xlsx` data, cleans it and converts it to tidy format.

## Installation

You can install the development version of vcglR like so:

``` r
# install.packages("devtools")
devtools::install_github("zerogetsamgow/vcglR")
```

## Gaming venue data


Data is stored as `egm_venue_data.rda` and `egm_lga_data.rda` allowing it to be loaded into R from the pacakge:

```{r}
egm_data = vcglR::egm_venue_data

dplyr::glimpse(egm_data)
```



Cleaning consists of:

-   updating and standardising venue names `venue_name` to the latest known venue name

-   standardising local government areas `lga_name` to an abbreviated format.

-   gathering quantitative data (the number of gaming machines and player loss) to long format and replacing missing data (0 counts of gaming machines). The type of data is denoted by `measure_type`, values are stored as `value`.

-   removing half-year data - only full financial year data is retained

-   cleaning the date information for each financial year and saving it as a character variable `financial_year` and date `fy_date` variable.

The package also adds geo-location data for each venue in the dataset. Location data was initially sourced from the regulator's [mapping tool](https://geomaps.vgccc.vic.gov.au/). Missing data was added manually and saved in `egm_locatons.csv` in `/data-raw/`

## Monthly LGA data

The Victorian gambling regulator publishes gaming machine data by LGA [on its website.](https://www.vgccc.vic.gov.au/resources/information-and-data/expenditure-data)

Released monthly, this dataset includes gaming expenditure for each LGA, and the number of gaming machines and venues operating in the LGA.

The data is published as `.xlsx` data. Historical data is also available.

The `vcglR` package reads this data, cleans it and converts it tidy format. Data is stored as `egm_lga_data.rda` allowing it to be loaded with:

```{r}
lga_data = vcglR::egm_lga_data

dplyr::glimpse(lga_data)
```

Cleaning consists of:

-   standardising local government areas `lga_name` to an abbreviated format.

-   gathering quantitative data (the number of gaming machines and player loss) to long format. The type of data is denoted by `measure_type`, values are stored as `value`.

-   cleaning the date information for each financial year and saving it as a date variable `financial_year` and month of data `data_month` variable.


## Last updated

Last update on 28 July 2025. Data for the financial year 2024-25 is included in the package datasets.
