---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 220
)

library(sf)
```

# vcglR

The Victorian gambling regulator publishes gaming machine data by venue and local government area [on its website.](https://www.vgccc.vic.gov.au/resources/information-and-data/expenditure-data)

Released twice a year for venue data and monthly for LGA data, these datasets include gaming expenditure at each gaming venue/local area, and the number of gaming machines it has.

The data is published as `.xlsx` data. Historical data is also available.

The goal of vcglR is to make Victorian gambling data available in tidy format. The name is taken from the abbreviation of the gambling regulators former name.

The `vcglR` package reads the `.xlsx` data, cleans it and converts it to tidy format.

## Installation

You can install `vcglR` like so:

``` r
# install.packages("remotes")
remotes::install_github("zerogetsamgow/vcglR")
```

## Gaming venue data

Data is stored as `egm_venue_data.rda` and `egm_lga_data.rda` allowing it to be loaded into R from the pacakge:

```{r}
egm_data = vcglR::egm_venue_data

dplyr::glimpse(egm_data)
```

Cleaning consists of:

-   updating and standardising venue names `venue_name` to the latest known venue name

-   standardising local government areas `lga_name` to an abbreviated format.

-   gathering quantitative data (the number of gaming machines and player loss) to long format and replacing missing data (0 counts of gaming machines). The type of data is denoted by `measure_type`, values are stored as `value`.

-   removing half-year data - only full financial year data is retained

-   cleaning the date information for each financial year and saving it as a character variable `financial_year` and date `fy_date` variable.

The package also adds geo-location data for each venue in the dataset. Location data was initially sourced from the regulator's [mapping tool](https://geomaps.vgccc.vic.gov.au/). Missing data was added manually and saved in `egm_locatons.csv` in `/data-raw/`

## Monthly LGA data

The Victorian gambling regulator publishes gaming machine data by LGA [on its website.](https://www.vgccc.vic.gov.au/resources/information-and-data/expenditure-data)

Released monthly, this dataset includes gaming expenditure for each LGA, and the number of gaming machines and venues operating in the LGA.

The data is published as `.xlsx` data. Historical data is also available.

The `vcglR` package reads this data, cleans it and converts it tidy format. Data is stored as `egm_lga_data.rda` allowing it to be loaded with:

```{r}
lga_data = vcglR::egm_lga_data

dplyr::glimpse(lga_data)
```

Cleaning consists of:

-   standardising local government areas `lga_name` to an abbreviated format.

-   gathering quantitative data (the number of gaming machines and player loss) to long format. The type of data is denoted by `measure_type`, values are stored as `value`.

-   cleaning the date information for each financial year and saving it as a date variable `financial_year` and month of data `data_month` variable.

## Last updated

Last update on `r format.Date(Sys.Date(),"%d %B %Y")`. Annual data is up to the `r fy::date2fy(Sys.Date()) |> fy::prev_fy()` financial year and monthly data is up to `r (lubridate::round_date(Sys.Date(), unit = "month") - lubridate::days(1) - lubridate::dmonths(1)) |> format.Date("%B %Y")`.

## Possible analysis

The following provides an example of some of the analysis possible with vcglR.

The Victorian Government has announced a trial of mandatory account-based play at dplyr::select venues from September. See <https://www.premier.vic.gov.au/new-trial-help-prevent-gambling-harm>

The trial will run across all 43 venues with gaming machines in Monash, Greater Dandenong and Ballarat from September to November.

`vcglR` can be used by help those who want to track the effects of the trial. The package includes `tidy` venue and LGA level gaming expenditure and machine count data. Venue locations are also geocoded to enable mapping and working with other data and geographic areas.

```{r analysis setup, include = TRUE}

venue_data = vcglR::egm_venue_data |> 
 # dplyr::filter(fy_date == max(fy_date)) |> 
  dplyr::select(
    tidyselect::contains("name"), 
    tidyselect::contains("type"), 
    value, lat, long, fy_date, financial_year) |> 
  dplyr::mutate(
    is_trial = 
      stringr::str_detect(
        lga_name, "Monash|Dandenong|Ballarat"))


lga_shapes =
  strayr::read_absmap(
    "lga2021", remove_year_suffix = TRUE) |> 
  dplyr::filter(state_code == 2) |> 
  dplyr::mutate(
    is_trial = 
      stringr::str_detect(
        lga_name, "Monash|Dandenong|Ballarat"))


```
Some early observations.

The trial in metropolitan LGAs should provide easily accessible data to measure the effect on gambling expenditure, noting we will be able to compare Monash and Dandenong with neighbouring LGAs. There are also some risks of players shifting to neighbouring venues to avoid the trial.

```{r make_melb_map, echo = FALSE}
metro_venue_map = 
  ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = lga_shapes, 
    mapping = 
      ggplot2::aes(
        geometry = geometry, 
        fill = is_trial),
    colour = vpstheme::bv.smoke) +
  ggplot2::geom_point(
    data = 
      venue_data |>
      dplyr::filter(fy_date == max(fy_date)) ,
    mapping = 
      ggplot2::aes(
        x = long,
        y = lat, 
        colour = is_trial)) +
  ggplot2::geom_text(
    data = lga_shapes, 
    size = 4, 
    mapping = 
      ggplot2::aes(
        label = stringr::str_wrap(lga_name,10), 
        x = cent_long, 
        y = cent_lat),
    colour = vpstheme::bv.charcoal) +
  ggplot2::scale_colour_manual(
    values =c( vpstheme::bv.navy, vpstheme::bv.purple),
    guide = "none") +
  ggplot2::scale_fill_manual(
    values =c( vpstheme::bv.white, vpstheme::bv.crystal), guide = "none") +
  ggplot2::coord_sf(
    ylim = c(-37.825,-38.1),
    xlim = c(145.0075,145.4)) +
  ggthemes::theme_map(); 

suppressWarnings(plot(metro_venue_map))

```

I will update the package as monthly data is released, so it will be possible to observe any effect on expenditure using the LGA level data as shown below.

```{r lga_expend_plot, echo = FALSE}
metro_lga_data =
  vcglR::egm_lga_data |> 
  dplyr::filter(
    stringr::str_detect(
      lga_name,
      "Casey|Dandenong|Monash|Eira|Knox|Whitehorse|Kingston")) |> 
  dplyr::filter(
    stringr::str_detect(measure_type, "Exp"), 
    !is.na(value)) 

rect =  data.frame(
  xmin = as.Date("2025-10-1"),
  xmax = as.Date("2025-12-1"),
  ymin = -Inf,  # Full vertical shading
  ymax = Inf
)

ggplot2::ggplot() +
  ggplot2::geom_rect(
    data = rect,
    inherit.aes = FALSE, 
    ggplot2::aes(
      xmin = xmin, 
      xmax = xmax, 
      ymin = ymin, 
      ymax = ymax),
      fill = vpstheme::bv.smoke)+
  ggplot2::geom_line(
    metro_lga_data,
    mapping = 
      ggplot2::aes(
        x = data_month,
        y = value, 
        colour = lga_name)
  ) +
  ggrepel::geom_text_repel(
    metro_lga_data |> 
      dplyr::filter(
        data_month == max(data_month)),
    hjust = 0,
    direction = "y",
    nudge_x = 10,
    lineheight = .7,
    mapping = 
      ggplot2::aes(
        x = data_month,
        y = value, 
        label = stringr::str_wrap(lga_name,10),
        colour = lga_name)
  ) +
  ggplot2::geom_vline(
    xintercept = 
      lubridate::ymd("2025-10-1"),
    colour = vpstheme::bv.charcoal,
    linetype = 2) +
  
  ggplot2::geom_text(
    data =
      tibble::tibble(
        "data_month" = lubridate::ymd("2025-10-10"),
        "value" = 1.5e6,
        "data_label" = "Trial\ncommenced"
      ),
    ggplot2::aes(
      x = data_month,
      y = value,
      label = data_label
    ),
    size = 4,
    lineheight = .7,
    hjust = 0,
    colour = vpstheme::bv.charcoal
  ) +
  vpstheme::scale_y_continuvps(
    name = "Net gambling expenditure, $ million",
    
    labels = scales::label_number(scale = 1/1e6)) +
  ggplot2::scale_x_date(
    name = NULL,
    date_labels = "%Y",
    breaks = seq.Date(
      from = lubridate::ymd("2023-6-1"), 
      to = lubridate::ymd("2026-6-1"), 
      by = "years"),
    expand = c(0,0)
  ) +
  ggplot2::coord_cartesian(
    xlim = lubridate::ymd("2023-8-1","2026-7-1"), 
    ylim = c(0,16e6)) +
  vpstheme::scale_colour_dffh(
    palette = "extended", 
    name = NULL, guide = "none")+
  vpstheme::theme_vps_djcs(base_size = 10)+
  ggplot2::theme(legend.position = "right")
```

Gaming machines in Ballarat have few substitutes, so players will find it more difficult to move to neighbouring venues to avoid the trial.

```{r make_ballarat_map, echo = FALSE}

ballarat_venue_map =
  ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = lga_shapes, 
    mapping = 
      ggplot2::aes(
        geometry = geometry, 
        fill = is_trial),
    colour = vpstheme::bv.smoke) +
  ggplot2::geom_point(
    data = venue_data,
    mapping = 
      ggplot2::aes(
        x = long,
        y = lat, 
        colour = is_trial)) +
  ggplot2::geom_text(
    data = lga_shapes, 
    size = 4, 
    mapping = 
      ggplot2::aes(
        label = stringr::str_wrap(lga_name,10), 
        x = cent_long, 
        y = cent_lat),
    colour = vpstheme::bv.charcoal) +
  ggplot2::scale_colour_manual(
    values =c( vpstheme::bv.navy, vpstheme::bv.purple), 
    guide = "none") +
  ggplot2::scale_fill_manual(
    values =c( vpstheme::bv.white, vpstheme::bv.crystal), 
    guide = "none") +
  ggplot2::coord_sf(
    ylim = c(-37.15,-37.9), 
    xlim = c(143.1,144.3)) +
  ggthemes::theme_map()

suppressWarnings(plot(ballarat_venue_map))


```

This means comparisons for Ballarat need to be made against similar regional centres rather than neighbouring LGAs. For example.

```{r regional_expend_plot, echo = FALSE}

regional_lga_data =
  vcglR::egm_lga_data |> 
  dplyr::filter(
    stringr::str_detect(
      lga_name,
      "Ballarat|Bendigo|Latrobe|Wodon")) |> 
  dplyr::filter(
    stringr::str_detect(measure_type, "Exp"), 
    !is.na(value)) 


ggplot2::ggplot() +
  ggplot2::geom_rect(
    data = rect,
    inherit.aes = FALSE, 
    ggplot2::aes(
      xmin = xmin, 
      xmax = xmax, 
      ymin = ymin, 
      ymax = ymax),
    fill = vpstheme::bv.smoke)+
 ggplot2:: geom_line(
    regional_lga_data,
    mapping = 
      ggplot2::aes(
        x=data_month,
        y = value, 
        colour = lga_name)
  ) +
  ggrepel::geom_text_repel(
    regional_lga_data |> 
      dplyr::filter(data_month == max(data_month)),
    hjust = 0,
    direction = "y",
    nudge_x = 10,
    #xlim =  lubridate::ymd("2025-5-31","2025-12-31"),
    lineheight = .4,
    mapping = 
      ggplot2::aes(
        x=data_month,
        y = value, 
        label = stringr::str_wrap(lga_name,10),
        colour = lga_name)
  ) +
  ggplot2::geom_vline(
    xintercept = lubridate::ymd("2025-10-1"),
    colour = vpstheme::bv.charcoal,
    linetype = 2) +
  ggplot2::geom_text(
    data =
      tibble::tibble(
        "data_month" = lubridate::ymd("2025-10-10"),
        "value" = 1.5e6,
        "data_label" = "Trial\ncommenced"
      ),
    ggplot2::aes(
      x = data_month,
      y = value,
      label = data_label
    ),
    size = 4,
    lineheight = .4,
    hjust = 0,
    colour = vpstheme::bv.charcoal
  ) +
  vpstheme::scale_y_continuvps(
    name = "Net gambling expenditure, $ million",
    labels = scales::label_number(scale = 1/1e6)) +
  ggplot2::scale_x_date(
    name = NULL,
    date_labels = "%Y",
    breaks = 
      seq.Date(
        from = lubridate::ymd("2023-6-1"), 
        to = lubridate::ymd("2026-6-1"), 
        by =   "years"),
    expand = c(0,0)
  ) +
  ggplot2::coord_cartesian(
    xlim = 
      lubridate::ymd(
        "2023-8-1",
        "2026-7-1"), 
    ylim = c(0,8e6)) +
  vpstheme::scale_colour_djcs(palette = "primary", name = NULL, guide = "none")+
  vpstheme::theme_vps_djcs(base_size = 10)+
  ggplot2::theme(legend.position = "right")
```

As venue data in the package is geocoded it supports analysis using other geographies like Statistical Areas.

```{r join_population, echo = FALSE}

sa3_shape =
  strayr::read_absmap(
    "sa32021", remove_year_suffix = TRUE) |> 
  dplyr::filter(state_code == 2)


venue_crs = 
  venue_data |> 
  sf::st_as_sf(
    coords = c("long", "lat"),
    crs = sf::st_crs(sa3_shape))

venue_sa3 = 
  sf::st_join(
    venue_crs,
    sa3_shape
  ) |> 
  tibble::as_tibble() |> 
  dplyr::left_join(
    venue_data, 
    by = 
      dplyr::join_by(
        venue_name, lga_name, venue_type,
        measure_type, value, fy_date, 
        financial_year, is_trial)
    ) |> 
  dplyr::select(
    tidyselect::contains("venue"),
    fy_date,
    is_trial,
    lat,
    long,
    measure_type, 
    value,
    tidyselect::contains("sa3"),
    ) |> 
  dplyr::group_by(sa3_name) |> 
  dplyr::mutate(sa3_trial = any(is_trial))


sa3_shape =
  dplyr::left_join(
    sa3_shape,
    venue_sa3 |>
      dplyr::filter(fy_date == max(fy_date)) |> 
      dplyr::select(contains("sa3")),
    by = dplyr::join_by(
      sa3_code, 
      sa3_name)) |> 
  dplyr::mutate(
    sa3_trial = 
      dplyr::coalesce(sa3_trial,FALSE))


```

```{r make_melb_sa3_map, echo = FALSE}
metro_sa3_map = 
  ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = sa3_shape, 
    mapping = 
      ggplot2::aes(
        geometry = geometry, 
        fill = sa3_trial),
    colour = vpstheme::bv.smoke) +
  ggplot2::geom_point(
    data = venue_sa3 |> 
      dplyr::filter(fy_date == max(fy_date)) ,
    mapping = 
      ggplot2::aes(
        x = long,
        y = lat, 
        colour = is_trial)
  ) +
  ggplot2::geom_text(
    data = sa3_shape, 
    size = 4, 
    mapping = 
      ggplot2::aes(
        label = stringr::str_wrap(sa3_name,10), 
        x = cent_long, 
        y = cent_lat),
    colour = vpstheme::bv.charcoal) +
  ggplot2::scale_colour_manual(
    values =c( vpstheme::bv.navy, vpstheme::bv.purple), guide = "none") +
  ggplot2::scale_fill_manual(
    values =c( vpstheme::bv.white, vpstheme::bv.crystal), guide = "none") +
  ggplot2::coord_sf(ylim = c(-37.825,-38.1), xlim = c(145.0075,145.4)) +
  ggplot2::labs(
    caption = 
      "Note: This chart use Statistical Areas Level 3 (SA3) geographical areas.") +
  ggthemes::theme_map(); 

suppressWarnings(plot(metro_sa3_map))

```

```{r get_pop, include = FALSE, echo = FALSE}

population =
  "https://www.abs.gov.au/statistics/people/population/regional-population/latest-release" |> 
  rvest::read_html() |> 
  rvest::html_elements('div [href$=xlsx]') |>
  rvest::html_attr("href")  |>
  tibble::as_tibble_col("url") |>
  dplyr::mutate(
    url = stringr::str_c("https://www.abs.gov.au", url)
  ) |>
  dplyr::filter(stringr::str_detect(url,"32180DS0003")) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    download = tempfile(fileext = ".xlsx")) |>
  dplyr::ungroup()  |>
  dplyr::mutate(
    x =
      purrr::pmap(
        list(url,download),
        function(a,b) if(!file.exists(b)) download.file(a,b, mode = "wb")))  |>
  dplyr::select(-x) |>
  dplyr::mutate(
    sheets = 
      purrr::map(download,readxl::excel_sheets)) |>
  tidyr::unnest(sheets) |>
  dplyr::filter(stringr::str_detect(sheets,"Table 2")) |>
  # Extract data
  dplyr::mutate(
    data = 
      purrr::pmap(
        list(download,sheets), readxl::read_excel, skip=5))  |>
  dplyr::select(data) |>
  tidyr::unnest(data) |> 
  dplyr::rename(
    "state_code" = 1,
    "state_name" = 2,
    "sa3_code" = 7,
    "sa3_name" = 8
  ) |> 
  dplyr::filter(state_code == 2) |> 
  dplyr::select(
    tidyselect::contains("sa3"),
    tidyselect::contains("20")) |> 
  tidyr::pivot_longer(
    tidyselect::contains("20"), 
    values_to = "population") |>
  dplyr::mutate(
    population = as.numeric(population),
    sa3_code = as.character(sa3_code),
    fy_date = stringr::str_c(name,"-6-30") |>
      lubridate::ymd())

```

```{r use_pop, echo = FALSE}
sa3_values_pop = 
  venue_sa3 |> 
  dplyr::left_join(
    population, 
    by = dplyr::join_by(
      fy_date, 
      sa3_code, 
      sa3_name)) |> 
  dplyr::group_by(
    fy_date, 
    sa3_name, 
    measure_type, 
    sa3_trial) |> 
  dplyr::summarise(
    population = dplyr::first(population),
    value = sum(value),
    .groups = "drop") |> 
  dplyr::mutate(
    per_1k = 1e3*value/population
  )

sa3_values_plot_data =
  sa3_values_pop |> 
  dplyr::ungroup() |> 
  dplyr::filter(!is.na(population)) |> 
  dplyr::filter(fy_date %in% c(max(fy_date))) |>
  dplyr::select(fy_date, sa3_name, measure_type, per_1k, sa3_trial) |> 
  tidyr::pivot_wider(
    values_from = per_1k, 
    names_from = measure_type)

sa3_plot = 
  ggplot2::ggplot() +
  ggrepel::geom_text_repel(
      sa3_values_plot_data |> 
        dplyr::filter(
          stringr::str_detect(
            sa3_name, 
            "Ballarat|Monash|Dandenong")),
      colour = vpstheme::bv.charcoal,
      ylim = c(.9e6,1e6),
      direction = "x",
      size = 5,
      segment.color = vpstheme::bv.smoke,
   mapping =
     ggplot2::aes(
       x = EGM,
       y = Expenditure,
       colour = sa3_trial,
       label = sa3_name))+
  ggplot2::geom_point(
    sa3_values_plot_data,
    mapping = 
      ggplot2::aes(
        x = EGM,
        y = Expenditure,
        colour = sa3_trial,
        group = sa3_name)) +
  vpstheme::scale_x_continuvps(
    name = "Gaming machines per 1000 people",
    limits = c(0, 11),
    breaks = seq(0,10,by=2)
  ) +
  vpstheme::scale_y_continuvps(
    name = 
      "Net gambling expenditure\nper 1000 people, $ 000s",
    limits = c(0, 1e6),
    labels = 
      scales:::label_number(scale = 1/1e3)) +
  ggplot2::scale_color_manual(
    guide = "none", 
    values =c( vpstheme::bv.purple, vpstheme::bv.navy)) +
  vpstheme::theme_vps_djcs(base_size = 10) +
  ggplot2::labs(
    caption = 
      "Note: This chart uses (SA3) geographical areas.")

suppressWarnings(plot(sa3_plot))
  



```
