---
title: "Density"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 200
)
library(sf)
```

```{r setup}
library(vcglR)
```

```{r get_pop, include = FALSE, echo = FALSE}

population =
  "https://www.abs.gov.au/statistics/people/population/regional-population/latest-release" |> 
  rvest::read_html() |> 
  rvest::html_elements('div [href$=xlsx]') |>
  rvest::html_attr("href")  |>
  tibble::as_tibble_col("url") |>
  dplyr::mutate(
    url = stringr::str_c("https://www.abs.gov.au", url)
  ) |>
  dplyr::filter(stringr::str_detect(url,"32180DS0003")) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    download = tempfile(fileext = ".xlsx")) |>
  dplyr::ungroup()  |>
  dplyr::mutate(
    x =
      purrr::pmap(
        list(url,download),
        function(a,b) 
          if(!file.exists(b)) {
            download.file(a,b,mode = "wb")})) |>
  dplyr::select(-x) |>
  dplyr::mutate(
    sheets = purrr::map(download,readxl::excel_sheets)) |>
  tidyr::unnest(sheets) |>
  dplyr::filter(stringr::str_detect(sheets,"Table 2")) |>
  # Extract data
  dplyr::mutate(
    data = 
      purrr::pmap(
        list(download,sheets), 
        readxl::read_excel, 
        skip = 5))  |>
  dplyr::select(data) |>
  tidyr::unnest(data) |> 
  dplyr::rename(
    "state_code" = 1,
    "state_name" = 2,
    "sa3_code" = 7,
    "sa3_name" = 8
  ) |> 
  dplyr::filter(state_code == 2) |> 
  dplyr::select(contains("sa3"),contains("20")) |> 
  tidyr::pivot_longer(tidyselect::contains("20"), values_to = "population") |>
  dplyr::mutate(population = as.numeric(population),
         fy_date = 
           stringr::str_c(name,"-6-30") |>
           lubridate::ymd())

venue_data = vcglR::egm_venue_data  |> 
  dplyr::mutate(
    is_trial = stringr::str_detect(lga_name, "Monash|Dandenong|Ballarat"))

```


# Introduction

The number of gaming machines relative to population in a given geographic area is strongly correlated with expenditure on gaming machine play. This correlation holds across State, local government area and statistical area geographies in most research.[^1]

```{r join_population, echo = FALSE}

sa3_shape =
  strayr::read_absmap("sa32021", remove_year_suffix = TRUE) |> 
  dplyr::filter(state_code == 2)


venue_crs = 
  venue_data |> 
  sf::st_as_sf(
    coords = c("long", "lat"),
    crs = st_crs(sa3_shape))

venue_sa3 = 
  sf::st_join(
    venue_crs,
    sa3_shape
  ) |> 
  tibble::as_tibble() |> 
  dplyr::left_join(
    venue_data, 
    by = 
      dplyr::join_by(
        venue_name, lga_name, venue_type,
        measure_type, value, fy_date, financial_year, is_trial)
    ) |> 
  dplyr::select(
    tidyselect::contains("venue"),
    fy_date,
    is_trial,
    lat,
    long,
    measure_type, 
    value,
    tidyselect::contains("sa3"),
    ) |> 
  dplyr::group_by(sa3_name) |> 
  dplyr::mutate(sa3_trial = any(is_trial))


sa3_shape =
  dplyr::left_join(
    sa3_shape,
    venue_sa3 |>
      dplyr::filter(fy_date == max(fy_date)) |> 
      dplyr::select(tidyselect::contains("sa3")),
    by = 
      dplyr::join_by(sa3_code, sa3_name)
  ) |> 
  dplyr::mutate(sa3_trial = dplyr::coalesce(sa3_trial,FALSE))

```


This correlation can be seen in Figure 1 below, which shows gaming machine density and net gambling expenditure per person at the statistical area (SA3) level in Victoria. The three regions highlighted are that participated in the mandatory carded play trial.

## Figure 1
```{r use_pop, echo = FALSE}
sa3_values_pop = 
  venue_sa3 |> 
  dplyr::left_join(
    population, 
    by = 
      dplyr::join_by(fy_date, sa3_code, sa3_name)) |> 
  dplyr::group_by(
    fy_date, sa3_name, 
    measure_type, sa3_trial) |> 
  dplyr::summarise(
    population = dplyr::first(population),
    value = sum(value),
    .groups = "drop") |> 
  dplyr::mutate(
    per_1k = 1e3*value/population
  )

sa3_values_plot_data =
  sa3_values_pop |> 
  dplyr::ungroup() |> 
  dplyr::filter(!is.na(population)) |> 
  dplyr::filter(fy_date %in% c(max(fy_date))) |>
  dplyr::select(
    fy_date, sa3_name, measure_type, 
    per_1k, sa3_trial) |> 
  tidyr::pivot_wider(
    values_from = per_1k, names_from = measure_type)

ggplot2::ggplot() +
ggrepel::geom_text_repel(
  sa3_values_plot_data |> 
    dplyr::filter(
      stringr::str_detect(
        sa3_name, 
        "Ballarat|Monash|Dandenong")),
    colour = vpstheme::bv.charcoal,
    ylim = c(.9e6,1e6),
    direction = "x",
    size = 5,
    segment.color = vpstheme::bv.smoke,
  mapping =  
    ggplot2::aes(
      x = EGM,
      y = Expenditure,
      colour = sa3_trial,
      label = sa3_name))+
  ggplot2::geom_point(
    sa3_values_plot_data,
    mapping = 
      ggplot2::aes(
        x = EGM,
        y = Expenditure,
        colour = sa3_trial,
        group = sa3_name)) +
  ggplot2::coord_cartesian(
    xlim = c(0, 11), 
    ylim = c(0, 1e6)) +
  vpstheme::scale_x_continuvps(
    name = "Gaming machines per 1000 people",
    breaks = seq(0,10,by=2)
  ) +
  vpstheme::scale_y_continuvps(
    name = "Net gambling expenditure\nper 1000 people, $ 000s",  ,
    labels = scales:::label_number(scale = 1/1e3)) +
  ggplot2::scale_color_manual(
    guide = "none", 
    values = c(
      vpstheme::bv.purple, 
      vpstheme::bv.navy)) +
  vpstheme::theme_vps_djcs(base_size = 10) +
  ggplot2::labs(title = 
         stringr::str_c(
           "Gaming machines and net gambling ",
           "expenditure in Victoria at the SA3 level"))

``` 

However, this correlation is a reflection of decades of policital, commercial and regulatory decisions about the location of gaming machines, not to mention urban development generally.

If we isolate out the regions that participated in the mandatory carded play trial and explore the relationship of gaming machine density and net gambling expenditure per person over time, we can see that this correlation becomes more complicated. 

Each line in Figure 2 shows the relationship for each SA3 starting in 2008-09 (the labelled end of the line) and progressing through to 2023-24. Years affected by venue closures as part of the public health response to the COVID-19 pandemic have been excluded as expenditure in these years was principally driven by the number of days venues could open.

In Ballarat, we can see that declining gaming machine density over time had little effect on expenditure per person. Apart from a discontinuity midway through the period of observation, expenditure per person was largely stable, despite a gradual reduction in density over time. The discontinuity in the Ballarat data, which we will investigate further, was likely due to a venue closure or relocation.

In Dandenong, we can see the relationship held over the start of the period, however the period spanning the pandemic was associated with an increase in per person expenditure at a time density was declining. 

Together, the patterns observed at small scales and over time suggest the relationship between gaming machine density and expenditure is complex and worthy of further investigation.

## Figure 2
```{r trial_lines, echo = FALSE}
trial_lines_data =
  sa3_values_pop |> 
  dplyr::ungroup() |> 
  dplyr::filter(
    !is.na(population),
    !(lubridate::year(fy_date) %in% 2020:2022) )  |>
  dplyr::select(
    fy_date, sa3_name, measure_type, per_1k, sa3_trial) |> 
  tidyr::pivot_wider(
    values_from = per_1k, 
    names_from = measure_type) |> 
  dplyr::mutate(
    ratio = EGM/Expenditure
  ) |> 
  dplyr::group_by(sa3_name) |> 
  dplyr::mutate(
    change = dplyr::last(ratio) - 
      dplyr::first(ratio)
  ) 

ggplot2::ggplot() +
  ggplot2::geom_path(
    trial_lines_data |> 
      dplyr::filter(sa3_trial),
   mapping = 
     ggplot2::aes(
        x = EGM,
        y = Expenditure,
        colour = sa3_name,
        group = sa3_name)
   ) +
  ggrepel::geom_text_repel(
    trial_lines_data |> 
      dplyr::filter(sa3_trial,fy_date == min(fy_date)),
      colour = vpstheme::bv.charcoal,
      ylim = c(.9e6,1e6),
      direction = "x",
      size = 5,
      segment.color = vpstheme::bv.smoke,
  mapping =  
    ggplot2::aes(
      x = EGM,
      y = Expenditure,
      colour = sa3_name,
      label = sa3_name)) +
  ggplot2::coord_cartesian(
    xlim = c(0, 11), ylim = c(0, 1e6)) +
  vpstheme::scale_x_continuvps(
    name = "Gaming machines per 1000 people",
    breaks = seq(0,10,by=2)
  ) +
  vpstheme::scale_y_continuvps(
    name = 
      "Net gambling expenditure\nper 1000 people, $ 000s",
    labels = scales:::label_number(scale = 1/1e3)) +
  ggplot2::scale_color_manual(
    guide = "none", 
    values = 
      c(
        vpstheme::bv.purple, 
        vpstheme::bv.cerulean, 
        vpstheme::bv.charcoal)) +
  vpstheme::theme_vps_djcs(base_size = 10) +
  ggplot2::labs(
    title = 
      stringr::str_c(
        "Gaming machines and net gambling ",
        "expenditure overtime in select Victorian SA3s"))
```

A manual inspection of the data shows there are several regions that exhibited increases in expenditure per person that confound the general relationship. These regions exhibited increases in gambling expenditure of more than $100 per person with negligible change in density. 


```{r exception_lines, echo = FALSE}

exception_lines_data =
  sa3_values_pop |> 
  dplyr::ungroup() |> 
  dplyr::filter(!is.na(population),
         !(lubridate::year(fy_date) %in% 2020:2022) ) |> 
  dplyr::mutate(
    other_trial = 
      stringr::str_detect(
        sa3_name, 
        stringr::str_flatten(
          c("Mildura",
            "Brimbank",
            "Shep",
            "Gippsland - East"),
          collapse = "|"))) |> 
  dplyr::filter(other_trial) |>
  dplyr::select(
    fy_date, sa3_name, measure_type, per_1k, sa3_trial) |> 
  tidyr::pivot_wider(
    values_from = per_1k, 
    names_from = measure_type) |> 
  dplyr::mutate(
    ratio = EGM/Expenditure
  ) |> 
  dplyr::group_by(sa3_name) |> 
  dplyr::mutate(
    change = ratio - dplyr::lag(ratio)
  )
  
# Make plot
ggplot2::ggplot() +
  ggplot2::geom_path(
   exception_lines_data,
   mapping = 
     ggplot2::aes(
        x = EGM,
        y = Expenditure,
        colour = sa3_name,
        group = sa3_name)
   ) +
  ggrepel::geom_text_repel(
   exception_lines_data |> 
      dplyr::filter(fy_date == min(fy_date)),
   colour = vpstheme::bv.charcoal,
    ylim = c(.9e6,1e6),
   direction = "x",
   size = 5,
   segment.color = vpstheme::bv.smoke,
  mapping =  
    ggplot2::aes(
      x = EGM,
      y = Expenditure,
      colour = sa3_name,
      label = sa3_name)) +
  ggplot2::coord_cartesian(
    xlim = c(0, 11), ylim = c(0, 1e6)) +
  vpstheme::scale_x_continuvps(
    name = "Gaming machines per 1000 people",
    breaks = seq(0,10,by=2)
  ) +
  vpstheme::scale_y_continuvps(
    name = 
      "Net gambling expenditure\nper 1000 people, $ 000s",  
    labels = scales:::label_number(scale = 1/1e3)) +
  ggplot2::scale_colour_manual(
    guide = "none", 
    values = 
      unname(vpstheme::bv_colours$primary)) +
  vpstheme::theme_vps_djcs(base_size = 10) +
  ggplot2::labs(title = 
         stringr::str_c(
           "Gaming machines and net gambling ",
           "expenditure overtime in select Victorian SA3s"))
```

[^1]: [Turning Point, 2015, Problem Gambling in Victoria: Identifying local area community and gaming industry risk and protective factors](https://responsiblegambling.vic.gov.au/documents/95/Research-report-pg-in-vic-community-industry-risk-protective-factors.pdf)
